\section{Exercises}\label{sec:11-path-algebras:06-exercises:1}

\textbf{Key points.} A quiver is just vertices, arrows, and source/target functions; a path is an inductive type indexed by its own endpoints, so an ill-typed composition (arrow source ≠ path's current endpoint) is rejected before any proof is attempted. \texttt{Path.append} composes two paths by recursion on the second, and the path algebra \(kQ\) --- sketched but not fully built here --- is the free \(k\)-module on the set of paths, with multiplication by composition.

\textbf{Socratic questions.}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \emph{\texttt{Quiver.Path}'s composability check (an arrow's source must match the path's current endpoint) is enforced by Lean at compile time. What would the equivalent check look like in an ordinary, non-dependently typed program, and when would it fire?} At runtime, as an \texttt{if}-check or assertion inside whatever function tries to append --- discovered the moment that specific composition executes, not before. Here, an ill-typed \texttt{Path.cons} simply has no proof term to supply, so the program that would have contained the bug never compiles at all.
\item
  \emph{A quiver with a genuine cycle (Exercise 1's \texttt{gamma}) has infinitely many distinct paths (go around the loop any number of times). Does that mean \texttt{Path Q u v} is an infinite }type\emph{, for such a \texttt{Q}?} For at least one pair \texttt{u = v} on the cycle, yes --- infinitely many distinct terms inhabit that one type, one per lap. This is exactly why a quiver's path algebra \(kQ\) is typically infinite-dimensional once it has a cycle, unless further relations are imposed.
\item
  \emph{The path algebra \(kQ\) is described as ``the free \(k\)-module on the set of paths,'' but this chapter builds only the paths and their composition, not \(kQ\) itself. What is the one piece of extra machinery still missing?} Finitely-supported functions from paths to \(k\) --- the ability to form a \(k\)-linear combination of \emph{finitely many} paths at once, which is genuinely more than what an inductive \texttt{Path} type alone provides.
\item
  Add a third arrow \texttt{gamma : ExampleArrow} with \texttt{source gamma = 2} and \texttt{target gamma = 0}, creating a cycle \texttt{0 → 1 → 2 → 0}. Build the path \texttt{gamma ∘ beta ∘ alpha : Path exampleQuiver 0 0}.
\item
  Prove \texttt{theorem append\_\allowbreak{}nil\_\allowbreak{}left \{V A\} \{Q : Quiver V A\} \{u v\} (p : Path Q u v) : Path.append (Path.nil u) p = p} by induction on \texttt{p} (mirroring the structure of \texttt{Path.append}'s own recursion), spelling out the base (\texttt{Path.nil}) and inductive (\texttt{Path.cons}) cases as in Chapter 4's \texttt{induction} examples.
\item
  (Optional, harder) Sketch --- in comments, no need to complete the Lean --- what a \texttt{structure PathAlgebra (V A : Type) (Q : Quiver V A) (k : Type) (Rg : Ring k)} would need to contain to satisfy \texttt{Ring}'s fields from Chapter 8.
\end{enumerate}

Solutions: \hyperref[sec:14-appendix-solutions:10-chapter-11:1]{Appendix, Chapter 11}.

\section{Next}\label{sec:11-path-algebras:06-exercises:2}

Continue to the \hyperref[sec:11-path-algebras:07-checkpoint-project:1]{checkpoint project}, which closes out Part II before Chapter 12 begins Part III.
