\section{Structuring lemmas for reuse}\label{sec:12-working-efficiently:05-structuring-lemmas:1}

The single biggest efficiency gain, greater than any tactic choice, is to \textbf{prove the general fact once, as its own named lemma, as soon as an argument would otherwise be repeated.} Chapter 7's \texttt{left\_\allowbreak{}inverse\_\allowbreak{}unique} is the running example. Theorem 3 (\texttt{inv\_\allowbreak{}op}) and Chapter 9's \texttt{neg\_\allowbreak{}one\_\allowbreak{}mul} and \texttt{neg\_\allowbreak{}mul} all reduce to it, instead of re-deriving ``uniqueness of inverses'' inline. Signs that a lemma should be factored out:

\begin{itemize}
\tightlist
\item
  A \href{https://lean-lang.org/doc/reference/latest/Tactic-Proofs/Tactic-Reference/}{\texttt{rw}} chain already written for a different but structurally identical goal is about to be repeated. In that case, the shared shape should instead be stated as its own \texttt{theorem}/\texttt{have}, then applied via \texttt{apply}/\texttt{exact} in both places.
\item
  A sub-goal deep in a proof would itself be a reasonable, independently statable mathematical fact (for example, ``an element that equals its own double is zero,'' buried inside Chapter 9's \texttt{mul\_\allowbreak{}zero}). Naming it is both more efficient \emph{and} more readable, since the outer proof then reads as a short chain of named facts instead of one long undifferentiated block.
\end{itemize}

This is the same judgment call made writing ordinary code: extract a helper when --- and only when --- real duplication or a genuinely separate sub-claim is present, not ahead of time ``just in case.''

\textbf{Key points.} \texttt{exact?}/\texttt{apply?} search for a closing term but do not always find the shortest one; \texttt{decide}/\texttt{omega}/\texttt{norm\_\allowbreak{}num} replace a hand proof exactly on their decidable fragment, never on a goal about a generic, unspecified structure; \texttt{simp} trades traceability for speed, so this book reaches for it only when a genuine technical obstacle makes the explicit alternative not worth the detour; and a repeated \texttt{rw} chain or an independently-statable sub-goal is a lemma waiting to be named.

\textbf{Socratic questions.}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \emph{\texttt{exact?} sometimes returns a working but needlessly roundabout term instead of the shortest one (Chapter 12 §1's own worked example). Does that make it a poor tool to reach for?} No --- it still guarantees a \emph{correct} closing term, found automatically; simplifying what it returns by hand afterward is a small, separate step, cheaper than deriving the whole thing from nothing.
\item
  \emph{\texttt{decide} and a hand-written proof can close the exact same goal on a finite carrier like \texttt{Fin 3}. If the end result is identical, why does this book insist on the hand-written version through Chapter 11, only endorsing \texttt{decide} here in Chapter 12?} Because the \emph{goal} of those earlier chapters was building the skill of finding a proof, which a decision procedure bypasses entirely. Once that skill is in hand, reaching for \texttt{decide} on a goal that is genuinely inside its decidable fragment is not a shortcut around understanding --- it is applying understanding efficiently.
\item
  \emph{A repeated \texttt{rw} chain is a sign a lemma should be factored out. Is the reverse also true --- should every proof be split into the smallest possible named lemmas, just in case one gets reused later?} No --- speculative extraction with no present duplication is the same mistake in the opposite direction: three similar lines are better than a premature abstraction. The signal to extract is an actual repeated shape or an actual independently-statable sub-claim, not the mere possibility of future reuse.
\end{enumerate}

\section{Next}\label{sec:12-working-efficiently:05-structuring-lemmas:2}

Continue to \hyperref[sec:13-next-steps:00-index:1]{Chapter 13: Where to go next}.
