\section{Exercises}\label{sec:05-rigor-check:05-exercises:1}

\textbf{Key points.} This book delays \texttt{class} for \texttt{structure} so every proof obligation stays explicit; \texttt{class} only changes \emph{how} Lean finds an instance, not what data it holds. \texttt{Type}'s own type must live one universe up (\texttt{Type 1}), or \texttt{Type : Type} reintroduces Russell's paradox. \texttt{rfl} proves only \emph{definitional} equality --- reduction to the same normal form --- which is not every true propositional equality (an asymmetric recursion like \texttt{Nat.add}'s is exactly where the two can diverge).

\textbf{Socratic questions.}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \emph{\texttt{class} only changes how Lean finds an instance, not what data it holds. If that is true, why does Mathlib bother with the whole \texttt{class} hierarchy at all, instead of using plain \texttt{structure}s like this book does?} Because ``how it is found'' is exactly the ergonomic payoff a huge shared library needs: automatic instance resolution means \texttt{+}/\texttt{*}/\texttt{1} just work at every registered type with no \texttt{Grp : Group G} argument threaded through every theorem by hand. This book delays that payoff on purpose, so the underlying data is seen plainly first.
\item
  \emph{\texttt{Type : Type} would make Lean's own type system inconsistent. Given that, why does \texttt{Type 1 : Type 2}, \texttt{Type 2 : Type 3}, and so on not cause the exact same problem one level up?} Because the paradox needs a type containing \emph{itself} --- \texttt{Type i : Type (i+1)} is always one level higher than what it classifies, so no universe ever contains its own code. The hierarchy is infinite specifically so this containment can never close the loop.
\item
  \emph{\texttt{rfl} and \texttt{decide} both look like they ``just work'' without argument. Are they doing the same thing?} No --- \texttt{rfl} checks that two terms reduce to an identical normal form; \texttt{decide} runs a \texttt{Decidable} instance's algorithm to compute \texttt{true}/\texttt{false} outright. They overlap on simple closed arithmetic (both close \texttt{2 + 2 = 4}), but only \texttt{decide} can settle something like \texttt{¬ (3 ∣ 10)}, which is not a reduction question at all.
\item
  Predict, before running it, whether \href{https://lean-lang.org/doc/reference/latest/Tactic-Proofs/Tactic-Reference/}{\texttt{example : (2 : Nat) * 3 = 3 + 3 := rfl}} type-checks. Then predict whether \texttt{example (n : Nat) : n * 2 = n + n := rfl} type-checks (hint: which argument does \texttt{Nat.mul} recurse on? Compare with the \texttt{Nat.add} recursion pattern from Chapter 4).
\item
  Rewrite \texttt{opTwice} (from the \texttt{structure} vs \texttt{class} section) as a type class version yourself: declare \texttt{class MyGroup (G : Type) where ...} with the same fields as this book's \texttt{Group}, register \texttt{instance : MyGroup Int where ...}, and write \texttt{def opTwiceTC [MyGroup G] (x : G) : G := MyGroup.op x x}. Confirm \texttt{\#eval opTwiceTC (3 : Int)} works with no explicit instance argument.
\item
  In one or two sentences, explain why \texttt{Type → Type} (the type of \texttt{Group} itself, before applying it to a carrier) must live in \texttt{Type 1} rather than \texttt{Type 0} --- tie the answer back to the Russell's-paradox obstruction this chapter described.
\item
  Give an example (distinct from \texttt{my\_add\_comm}) of a true propositional equality between two \texttt{Nat} expressions that is \emph{not} provable by \texttt{rfl} alone, and identify which of the two sides' recursive structure is the obstruction.
\end{enumerate}

Solutions: \hyperref[sec:14-appendix-solutions:04-chapter-5:1]{Appendix, Chapter 5}.

\section{Next}\label{sec:05-rigor-check:05-exercises:2}

Continue to the \hyperref[sec:05-rigor-check:06-checkpoint-project:1]{checkpoint project}, which closes out Part I before Chapter 6 begins Part II.
