% ===================================================================
% Title page
% ===================================================================
\begin{titlepage}
\centering
\vspace*{2cm}

{\color{ruleblue}\rule{0.6\textwidth}{1.2pt}}\par
\vspace{0.6cm}

{\normalfont\Huge\bfseries Lean for Working Algebraists\par}
\vspace{0.5cm}
{\normalfont\Large\itshape A Guide to Formalizing Groups, Rings, and\\[2pt]
Path Algebras in Lean~4\par}

\vspace{0.6cm}
{\color{ruleblue}\rule{0.6\textwidth}{1.2pt}}\par

\vspace{2.5cm}
{\normalfont\Large Abderrahim Adrabi\par}

\vfill

{\normalfont\large\itshape "Every proof in this book was verified against
a real Lean~4 toolchain --- \\
not just described."\par}

\vspace{1.5cm}
{\normalfont\normalsize \today\par}
{\normalfont\small Lean toolchain \texttt{v4.31.0}\par}

\end{titlepage}

% ===================================================================
% Preface
% ===================================================================
\chapter*{Preface}

This is a guide to Lean~4 for readers who already think in groups, rings,
functors, and diagrams, but have never written a line of Lean --- or any
proof assistant. No programming background is assumed, only the
mathematical maturity of someone comfortable with abstract algebra and the
basic language of category theory: objects, morphisms, composition,
functors. The book builds up Lean~4 syntax and tactics only as far as
needed to formalize groups, rings, and quiver path algebras; the
categorical viewpoint --- a quiver's path category, a ring as a one-object
preadditive category, and so on --- is called out explicitly wherever it
clarifies what the Lean code is really encoding.

This book is, and remains, Mathlib-free by design through Chapter~5: every
group, ring, and path algebra is built from scratch, so every definition
and proof obligation stays visible. Starting in Chapter~6, each worked
example is followed by a ``Mathlib equivalent'' box showing the same
statement phrased against Mathlib's real API --- not a contradiction of
the from-scratch approach, but a second, parallel track. Seeing the same
idea twice, once built by hand and once as a library already has it, is
how both halves of working in Lean get learned at once: the mathematics
itself, and the shape of the library a reader will actually use
afterward. Chapter~13 completes the handoff to Mathlib in full.

The running goal throughout is more than the constructs covered: it is
the \emph{skill} of using Lean --- reading a goal state, deciding what to
try next, recovering when a tactic fails, and knowing which proofs to
derive by hand versus hand off to automation.

% ===================================================================
% About this book
% ===================================================================
\chapter*{About this book}
This book was written as an experiment in using Claude Code (Anthropic's
AI coding agent) to produce a complete technical book: the author,
Abderrahim Adrabi, directed the scope, structure, audience, and every
round of revision across many sessions; Claude wrote the chapter text,
Lean code, and LaTeX. No human Lean expert or mathematician has
independently reviewed the mathematical or Lean content beyond the
author's own direction and the automated checks described in the
project's README and CHANGELOG (every code block compiled against a real
Lean~4 toolchain, plus several AI-driven review passes). Treat it as a
self-checked draft produced through human-directed AI authorship, not a
traditionally peer-reviewed text.

% ===================================================================
% How to read this book
% ===================================================================
\chapter*{How to read this book}

Not every reader needs every chapter in order. The Learning paths chapter
just after this one lays out a chapter-dependency graph and a handful of
named paths through the book --- already know Lean? already know algebra?
want the formal foundations first? --- for readers who would rather not
start cover to cover.

Each chapter is divided into short, numbered sections building on one
another. Every code listing is valid Lean~4 (toolchain
\texttt{v4.31.0}). Every code block in Chapters~1--11 has additionally
been ported into a companion Lean project and verified with
\texttt{lake build} against the real compiler, not merely read over ---
several bugs that only surfaced this way (a missing extensionality
lemma, a couple of \texttt{rw} steps that over-rewrote, a reference to an
undefined name, a tactic used without its library ever being imported)
have been fixed in both the book and the project.

This book is about more than the constructs it covers. Chapters~7 and~9
present each theorem as a search process --- what to look at, what to
try, why an attempt fails --- rather than only the polished final proof;
Chapter~12 is dedicated entirely to working efficiently once the
underlying ideas are understood. Chapter~5 pauses to address the rigor
questions a careful mathematician will already be asking by that point
--- \texttt{structure} versus \texttt{class}, universes, and definitional
versus propositional equality --- before Chapter~6 commits to
\texttt{Group}'s definition.

A notation reference follows this section, for quick lookup as unfamiliar
symbols are encountered; two further lookup tables --- a tactic and
library reference, and a $\lambda$-calculus-to-Lean dictionary --- sit at
the back of the book, after the last chapter, to be consulted as needed
rather than read start to finish.

\tableofcontents
